---
title: "1.1_data_preprocessing_CamCAN"
output: html_document
date: "2025-08-30"
---

```{r}
library(tidyverse)
library(lmerTest)
library(gamm4)
```

# Longitudinal data preparation 
## Cognitive Data Preprocessing 
```{r, include=TRUE}
ct_data  <- read.csv("data_raw/longitudinal/cog_data/CamCAN_Cattell_REDCAP_PH245_06_09_2025.csv") 
ppt      <- read.csv("data_raw/longitudinal/cog_data/CamCAN_PPT_ALL_06_09_2025.csv")        
frails   <- read.csv("data_raw/longitudinal/cog_data/CamCAN_FRAIL_CCID_06_09_2025.csv")           
edu      <- read.csv("data_raw/longitudinal/cog_data/CamCAN_YoE_06_09_2025.csv")
```

```{r, include=TRUE}
ct_data$arm   <-  ifelse(grepl("_arm_1", ct_data$redcap_event_name),  "1", 
                  ifelse(grepl("_arm_2", ct_data$redcap_event_name),  "2", NA))

ct_data$phase <-  ifelse(grepl("phase_5", ct_data$redcap_event_name), "5", 
                  ifelse(grepl("phase_4", ct_data$redcap_event_name), "4",
                  ifelse(grepl("phase_2", ct_data$redcap_event_name), "2", NA)))

ct_data$wave  <-  ifelse(grepl("phase_2", ct_data$redcap_event_name), "0", 
                  ifelse(grepl("phase_4", ct_data$redcap_event_name), "1",
                  ifelse(grepl("phase_5", ct_data$redcap_event_name), "2", NA)))

ct_data <- merge(ppt, ct_data, by = "record_id")
ct_data <- merge(ct_data, edu, by = "record_id")

ct_data <- ct_data %>% dplyr::select(-c("handedness", "phase_cog_age", "redcap_event_name",
                                        "redcap_event_name.x", "redcap_event_name.y"))

names(ct_data)[names(ct_data) == "record_id"]              <- "CC.ID"
names(ct_data)[names(ct_data) == "sex"]                    <- "Sex"
names(ct_data)[names(ct_data) == "phase_code"]             <- "frail_code"
names(ct_data)[names(ct_data) == "phase_cog_online_paper"] <- "Format"
names(ct_data)[names(ct_data) == "cattell_total_score"]    <- "Cattell"
names(ct_data)[names(ct_data) == "question_num_years_edu"] <- "YoE"

# names(frails)[names(frails) == "record_id"]              <- "CC.ID"

ct_data$dob          <- as.Date(ct_data$dob)
ct_data$cattell_date <- as.Date(ct_data$cattell_date)
ct_data$Age          <- round(as.numeric((ct_data$cattell_date - ct_data$dob) / 365.25), 2)

ct_data <- ct_data[, c("CC.ID", "phase", "wave", "Age", "Sex", "YoE",
                       "Cattell", "Format",     
                       "cattell_subscore_1","cattell_subscore_2", 
                       "cattell_subscore_3","cattell_subscore_4",
                       "arm", "dob", "cattell_date", "frail_code")]

ct_data$CC.ID   <- as.character(ct_data$CC.ID)

ct_data$phase   <- as.factor(ct_data$phase)
ct_data$wave    <- as.factor(ct_data$wave)
ct_data$Format  <- as.factor(ct_data$Format)
ct_data$Sex     <- as.factor(ct_data$Sex)

ct_data$Age     <- as.numeric(ct_data$Age)
ct_data$YoE     <- as.numeric(ct_data$YoE)
ct_data$Cattell <- as.numeric(ct_data$Cattell)

# Subject pattern grouping:  

# G1: [100]	          cc700 only 
# G2: [110]           cc700+ccEnrich (rescan dropput)
# G3: [111] or [101]	cc700+ccEnrich+ccrescan or cc700+ccrescan (true 'returners')

# G3 - longitudinal target group
# Note: NAs are assigned to 010 (Enrich only)

subject_patterns <- ct_data %>%
  mutate(
    participated = ifelse(!is.na(Cattell), 1, 0)
  ) %>%
  group_by(CC.ID) %>%
  summarize(
    wave0 = max(ifelse(wave == 0, participated, 0)),
    wave1 = max(ifelse(wave == 1, participated, 0)),
    wave2 = max(ifelse(wave == 2, participated, 0)),
  ) %>%
  mutate(
    pattern = paste0(wave0, wave1, wave2),
    Group = case_when(
      pattern == "100" ~ "G1",
      pattern == "110" ~ "G2",
      pattern %in% c("111", "101") ~ "G3",
      TRUE ~ NA_character_
    )
  )
subject_patterns <- subject_patterns[!is.na(subject_patterns$Group), ]

ct_data <- merge(ct_data, subject_patterns, by = "CC.ID")
ct_data <- ct_data[order(ct_data$CC.ID, ct_data$wave), ]

ct_data$Group <- as.factor(ct_data$Group)
ct_data$Practice                                <- 1 
ct_data$Practice[str_detect(ct_data$wave, "0")] <- 0
ct_data$Practice                                <- as.factor(ct_data$Practice)

ct_data <- ct_data %>%
  group_by(CC.ID) %>%
  mutate(A0 = as.numeric(Age[wave == '0']), 
         dA = as.numeric(Age - A0))

cols <- c("Cattell", "YoE", "Age", "A0", "dA", "Practice", "Format", "Sex")
for (col in cols) {
  ct_data[[paste0(col, "_raw")]] <- ct_data[[col]]
}

ct_data$Sex      <- as.numeric(ct_data$Sex)
ct_data$Practice <- as.numeric(ct_data$Practice)
ct_data$Format   <- as.numeric(ct_data$Format)

ct_data[cols] <- lapply(ct_data[cols], scale)

#OPTIONAL - exclude Frails (code 6) from the derivatives
# ct_data <- ct_data[!ct_data$CC.ID %in% frails$CC.ID, ]
```

### Adjustment for Practice and Format effects 
```{r, include=TRUE}
ct_data_clean <- ct_data[!is.na(ct_data$dA) & !is.na(ct_data$Cattell) & !is.na(ct_data$Format), ]
ct_data_G3    <- ct_data_clean[ct_data_clean$Group == "G3", ]
ct_data_G3    <- ct_data_G3[order(ct_data_G3$CC.ID, ct_data_G3$wave), ]

model1 <- lmer(Cattell_raw ~ Format_raw + Practice_raw + (1|CC.ID), data=ct_data_G3, REML=TRUE)

model2 <- lmer(Cattell_raw ~ A0:dA + Format_raw + Practice_raw + (1|CC.ID), data=ct_data_G3, REML=TRUE)

model3 <- gamm(Cattell_raw ~ s(Age_raw) + Format_raw + Practice_raw, data = ct_data_G3, random = list(CC.ID = ~1), REML   = TRUE)

ct_data_G3$Cattell_A0    <- residuals(model1)
ct_data_G3$Cattell_A0dA  <- residuals(model2)
ct_data_G3$Cattell_gamm  <- residuals(model3$gam, type = "response")
```

#Descriptive stats 
```{r}
length(unique(ct_data_G3$CC.ID))
length(unique(ct_data_G3$CC.ID[ct_data_G3$wave == 1 & !is.na(ct_data_G3$Cattell_raw)]))
```

### Cognitive Slopes calculation
```{r}
unique_ids        <- unique(ct_data_G3$CC.ID)
n                 <- length(unique_ids)
Cattell_slope_gamm     <- numeric(n)
Cattell_slope_A0dA     <- numeric(n)
Cattell_slope_A0     <- numeric(n)

Cattell0          <- numeric(n) 
Cattell1          <- numeric(n) 
Cattell2          <- numeric(n) 

for (i in seq_along(unique_ids)) {
  
  this_id          <- unique_ids[i]
  subset_data      <- ct_data_G3[ct_data_G3$CC.ID == this_id, ]
  
  model1           <- lm(Cattell_gamm ~ dA_raw, data = subset_data)
  Cattell_slope_gamm[i] <- coef(model1)[["dA_raw"]]
  
  model2           <- lm(Cattell_A0dA ~ dA_raw, data = subset_data)
  Cattell_slope_A0dA[i] <- coef(model2)[["dA_raw"]]
  
  model3           <- lm(Cattell_A0 ~ dA_raw, data = subset_data)
  Cattell_slope_A0[i] <- coef(model3)[["dA_raw"]]
  
  wave0 <- subset_data[subset_data$wave == 0, ]
  wave1 <- subset_data[subset_data$wave == 1, ]
  wave2 <- subset_data[subset_data$wave == 2, ]
  
  Cattell0[i]    <- wave0$Cattell[1]
  Cattell1[i]    <- wave1$Cattell[1]
  Cattell2[i]    <- wave2$Cattell[1]
  
}

cog_slopes <- data.frame(CC.ID         = unique_ids, 
                         Cattell_slope_gamm = Cattell_slope_gamm, 
                         Cattell_slope_A0dA = Cattell_slope_A0dA, 
                         Cattell_slope_A0   = Cattell_slope_A0, 
                         Cattell0      = Cattell0, 
                         Cattell1      = Cattell1, 
                         Cattell2      = Cattell2, 
                         stringsAsFactors = FALSE)

cols <- c("Cattell_slope_gamm", "Cattell_slope_A0dA", "Cattell_slope_A0", "Cattell0", "Cattell1", "Cattell2")
cog_slopes[paste0(cols, "_zscored")] <- lapply(cog_slopes[cols], scale)

write.csv(cog_slopes, "data_preprocessed/Cattell_dAs_adjusted_A0_prac_format.csv", row.names = FALSE)
```

## Neuroimaging data preprocessing (longitudinal) 
```{r}
mri_date <- read.csv("data_raw/longitudinal/MR_data/CambridgeCentreForAg-MRIDate_DATA_2025-07-14_2035.csv")
gm       <- read_tsv("data_raw/longitudinal/MR_data/aseg_vol_table.tsv")
lthick   <- read_tsv("data_raw/longitudinal/MR_data/lh.aparc.thickness.table.tsv")
rthick   <- read_tsv("data_raw/longitudinal/MR_data/rh.aparc.thickness.table.tsv")
larea    <- read_tsv("data_raw/longitudinal/MR_data/lh.aparc.area.table.tsv")
rarea    <- read_tsv("data_raw/longitudinal/MR_data/rh.aparc.area.table.tsv")

thickness <- right_join(lthick, rthick)
area      <- right_join(larea, rarea)
gm        <- merge(gm, thickness)
gm        <- merge(gm, area)

gm$record_id   <- sub(".*CC(\\d+).*", "\\1", gm$`Measure:volume`)
gm$CC.ID       <- gm$record_id
mri_date$CC.ID <- mri_date$record_id

names(gm)[names(gm) == "Measure:volume"] <- "id"
gm$phase <- as.numeric(gsub(".*_ses-P(\\d+).*", "\\1", gm$id))

gm$wave[gm$phase == 2] <- 0
gm$wave[gm$phase == 3] <- 1
gm$wave[gm$phase == 5] <- 2

gm <- merge(gm, ppt, by = c("record_id"))
gm <- merge(gm, ct_data[ct_data$wave == 0, c("CC.ID", "YoE_raw")], by = "CC.ID")

mri_date$wave <- ifelse(grepl("phase_2", mri_date$redcap_event_name), "0", 
                 ifelse(grepl("phase_3", mri_date$redcap_event_name), "1",
                 ifelse(grepl("phase_5", mri_date$redcap_event_name), "2", NA)))

gm <- merge(gm, mri_date, by = c("record_id", "wave"))

gm$dob      <- as.Date(gm$dob)
gm$mri_date <- as.Date(gm$phase_mri_date)
gm$MR_Age   <- round(as.numeric((gm$mri_date - gm$dob) / 365.25), 2)

gm$Thickness <- (gm$rh_MeanThickness_thickness + gm$lh_MeanThickness_thickness)/2
gm$Area      <- (gm$rh_WhiteSurfArea_area + gm$lh_WhiteSurfArea_area)/2
gm$WMV       <- (gm$rhCerebralWhiteMatterVol + gm$lhCerebralWhiteMatterVol)/2

gm$Thickness_z <- as.numeric(scale(gm$rh_MeanThickness_thickness + gm$lh_MeanThickness_thickness)/2)
gm$Area_z      <- as.numeric(scale(gm$rh_WhiteSurfArea_area + gm$lh_WhiteSurfArea_area)/2)
gm$WMV_z       <- as.numeric(scale(gm$rhCerebralWhiteMatterVol + gm$lhCerebralWhiteMatterVol)/2)

gm$CC.ID <- gm$CC.ID.x 
gm$CC.ID.x <- NULL
gm$CC.ID.y <- NULL
```

```{r}
all_vars <- names(gm)
lh_vars <- grep("^lh_.*_(thickness|area)$", all_vars, value = TRUE)
regions <- str_replace(lh_vars, "^lh_", "")

mean_vars <- list()
for (region in regions) {
  lh_col <- paste0("lh_", region)
  rh_col <- paste0("rh_", region)
  mean_col <- paste0("mean_", region)  
  if (lh_col %in% all_vars && rh_col %in% all_vars) {
    gm[[mean_col]] <- rowMeans(gm[, c(lh_col, rh_col)], na.rm = TRUE)
    mean_vars[[mean_col]] <- c(lh_col, rh_col)
  }
}

gm <- gm %>%
  group_by(CC.ID) %>%
  mutate(
    A0 = as.numeric(MR_Age[wave == '0']),
    dA = as.numeric(MR_Age - A0),
    MR_Age = as.numeric(MR_Age),
    Thickness_baseline = as.numeric(Thickness[wave == '0']),
    Area_baseline      = as.numeric(Area[wave == '0']), 
    WMV_baseline       = as.numeric(CerebralWhiteMatterVol[wave == '0']), 
    WMV_0              = as.numeric(WMV_z[wave == '0']), 
    WMV_2              = as.numeric(WMV_z[wave == '2']), 
    Thickness0         = as.numeric(Thickness_z[wave == '0']),
    Thickness2         = as.numeric(Thickness_z[wave == '2']),
    Area0              = as.numeric(Area_z[wave == '0']),
    Area2              = as.numeric(Area_z[wave == '2'])
  ) %>%
  ungroup()

adjustments <- list(
  A0 = function(var)
    as.formula(paste0(var, " ~ 1 + (1|CC.ID)")),
  A0dA = function(var)
    as.formula(paste0(var, " ~ A0:dA + (1|CC.ID)")),
  gamm = function(var)
    as.formula(paste0(var, " ~ s(MR_Age)"))
)

global_vars <- c("Thickness", "Area", "WMV")
mean_roi_vars <- names(gm)[grepl("^mean_.*_(thickness|area)$", names(gm))]
all_vars <- c(global_vars, mean_roi_vars)

for (v in all_vars) {
  ## A0
  m_A0 <- lmer(adjustments$A0(v), data = gm, REML = TRUE)
  gm[[paste0(v, "_A0")]] <- residuals(m_A0)
  ## A0 + A0:dA
  m_A0dA <- lmer(adjustments$A0dA(v), data = gm, REML = TRUE)
  gm[[paste0(v, "_A0dA")]] <- residuals(m_A0dA)
  ## GAMM (age)
  m_gamm <- gamm(formula = adjustments$gamm(v), data = gm, random  = list(CC.ID = ~1), REML = TRUE)
  gm[[paste0(v, "_gamm")]] <- residuals(m_gamm$gam, type = "response")
}
```

```{r}

roi_slopes <- data.frame(CC.ID = unique(gm$CC.ID))

for (v in all_vars) {
  for (adj in c("A0", "A0dA", "gamm")) {

    yvar <- paste0(v, "_", adj)
    slope_vec <- rep(NA_real_, nrow(roi_slopes))

    for (i in seq_along(roi_slopes$CC.ID)) {

      id <- roi_slopes$CC.ID[i]
      sub <- gm[gm$CC.ID == id, c("dA", yvar)]

      y <- as.numeric(sub[[yvar]])
      x <- as.numeric(sub$dA)

      if (sum(!is.na(y)) > 1 && var(x, na.rm = TRUE) > 0) {
        slope_vec[i] <- coef(lm(y ~ x))[2]
      }
    }

    roi_slopes[[paste0(v, "_", adj, "_slope")]] <- slope_vec
  }
}

# thk_slope_cols  <- grep("_thickness_slope$", names(roi_slopes), value = TRUE)
# area_slope_cols <- grep("_area_slope$", names(roi_slopes), value = TRUE)
# 
# roi_slopes$mean_thickness_slope <- rowMeans(roi_slopes[, thk_slope_cols], na.rm = TRUE)
# roi_slopes$mean_area_slope      <- rowMeans(roi_slopes[, area_slope_cols], na.rm = TRUE)

gm <- merge(gm, roi_slopes, by = "CC.ID", all.x = TRUE)

write.csv(gm, "data_preprocessed/CamCAN_longitudinal_GM_G3_4Sep25.csv", row.names = FALSE)
```

##Combine Cognitive and MR slopes 
```{r}
gm_slopes <- gm[gm$wave == 0, c("CC.ID", "A0", "sex", "YoE_raw", "Thickness_baseline", "Area_baseline", "Thickness0", "Thickness2", "Area0", "Area2",                                       "WMV_0", "WMV_2", "WMV_baseline",
                                grep("slope$", names(gm), value = TRUE))]
mydata_slopes <- merge(gm_slopes, cog_slopes, by = "CC.ID")
```

### Save longitudinal cognitive and neuroimaging data 
```{r}
write.csv(mydata_slopes, "data_preprocessed/CamCAN_longitudinal_slopes_Cattell_GM_30Aug25.csv", row.names = FALSE)
```

# Cross-sectional data preparation
## Read-in genetics data (Mains + Frails)
```{r}
IQ        <- read.csv("data_raw/cross_sectional/genetics/PRS_updated.csv")
IQ_frails <- read.csv("data_raw/cross_sectional/genetics/PRS_updated_frails.csv")
```

#Merge cognitive and genetic data 
```{r}
IQ    <- bind_rows(IQ, IQ_frails)
cc700 <- ct_data[ct_data$wave == 0,]

cc700$record_id <- cc700$CC.ID
IQ$record_id    <- IQ$CCID

cc700 <- merge(cc700, ppt, by = "record_id")
cc700 <- merge(cc700, IQ,  by = "record_id")
```

#Neuroimaging data 
```{r}
lthick700   <- read_table("data_raw/cross_sectional/MR_data/aparc_thickness_lh.txt")
rthick700   <- read_table("data_raw/cross_sectional/MR_data/aparc_thickness_rh.txt")
larea700    <- read_table("data_raw/cross_sectional/MR_data/aparc_area_lh.txt")
rarea700    <- read_table("data_raw/cross_sectional/MR_data/aparc_area_rh.txt")

lthick700_f <- read_table("data_raw/cross_sectional/MR_data/lh.aparc.thickness.table.frails.csv")
rthick700_f <- read_table("data_raw/cross_sectional/MR_data/rh.aparc.thickness.table.frails.csv")
larea700_f  <- read_table("data_raw/cross_sectional/MR_data/lh.aparc.area.table.frails.tsv")
rarea700_f  <- read_table("data_raw/cross_sectional/MR_data/rh.aparc.area.table.frails.tsv")
```

```{r}
thickness700      <- right_join(lthick700, rthick700)
area700           <- right_join(larea700, rarea700)
thickness_area700 <- right_join(thickness700, area700)

thickness_f       <- right_join(lthick700_f, rthick700_f)
area_f            <- right_join(larea700_f, rarea700_f)
thickness_area_f  <- right_join(thickness_f, area_f)

brain700            <- bind_rows(thickness_area700, thickness_area_f)
brain700$record_id  <- as.numeric(sub(".*(CC|PP)(\\d+)", "\\2", brain700$lh.aparc.thickness))
brain700$frail_flag <- ifelse(brain700$record_id %in% frails$CC.ID, 1, 0)
```

###Calculate mean thickness and area as an average of all regions 
```{r}
all_vars <- names(brain700)
lh_vars  <- grep("^lh_.*_(thickness|area)$", all_vars, value = TRUE)
regions  <- str_replace(lh_vars, "^lh_", "")

mean_vars <- list()

for (region in regions) {
  lh_col <- paste0("lh_", region)
  rh_col <- paste0("rh_", region)
  mean_col <- paste0("", region)
  
  if (lh_col %in% all_vars && rh_col %in% all_vars) {
    brain700[[mean_col]] <- rowMeans(brain700[, c(lh_col, rh_col)], na.rm = TRUE)
    mean_vars[[mean_col]] <- c(lh_col, rh_col)
  }
}

mydata700 <- merge(cc700, brain700, by = "record_id")

names(mydata700)[names(mydata700) == "MeanThickness_thickness"] <- "Thickness"
names(mydata700)[names(mydata700) == "WhiteSurfArea_area"]      <- "Area"
names(mydata700)[names(mydata700) == "IQ_SavageEtAl2018"]       <- "PGS"

mydata700$Thickness_raw <- mydata700$Thickness
mydata700$Area_raw      <- mydata700$Area 
mydata700$Thickness     <- scale(mydata700$Thickness_raw)
mydata700$Area          <- scale(mydata700$Area_raw)

mydata700$return_code <- ifelse(mydata700$Group == "G3", 1, 0)
```

### Save cross-sectional cognitive, neuroimaging and PGS data 
```{r}
write.csv(mydata700, "data_preprocessed/CamCAN_cross_sectional_Cattell_GM_PGS_3Sep25.csv", row.names = FALSE)
```
