---
always_allow_html: true
output:
  html_document: default
  word_document: default
  pdf_document: default
  always_allow_html: default
date: "2026-01-11"
title: "2.1_data_analysis_CamCAN"
---
```{r, include  = FALSE}
library(tidyverse) #dplyr, ...
library(car)       #linear hypothesis test
library(lmerTest)
library(lavaan)
library(cocor)     #Steiger's test
library(semptools) #SEM significance plotting 
library(gridExtra)
library(kableExtra)
library(corrplot)
library(semPlot)
library(tidySEM)
library(gridGraphics)
library(effectsize)
library(broom.mixed)
library(gt)

# options(scipen = 999)
```

### Custom functions for plotting
```{r, include = TRUE}
plotcross <- function(data, x_var, y_var, x_label, y_label, ymin, ymax, color_choice) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    geom_point(size = 2.5, stroke = 1, alpha = 1, color = color_choice) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 1, lty = 1) +
    # geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "black", linewidth = 1, lty = 1) +
    labs(x = x_label, y = y_label) +
    ylim(ymin, ymax) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.title.x = element_text(size = 24),
      axis.title.y = element_text(size = 24), 
      axis.text.x = element_text(size = 10),  
      axis.text.y = element_text(size = 10)  
    )
}
plotit <- function(data, x_var, y_var, x_label, y_label, ymin, ymax, color_choice) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    geom_point(size = 2.5, stroke = 1, alpha = 0.3, color = color_choice) +
    geom_line(aes(group = CC.ID), lty = 2, alpha = 0.1, color = "black") +
    geom_smooth(method = "lm", se = FALSE, color = color_choice, size = 0.5, lty = 1, aes(group = CC.ID)) +
    labs(x = x_label, y = y_label) +
    ylim(ymin, ymax) +
    theme_minimal() + 
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.title.x = element_text(size = 24),
      axis.title.y = element_text(size = 24), 
      axis.text.x = element_text(size = 16),  
      axis.text.y = element_text(size = 16)  
    )
}
plotit2 <- function(data, x_var, y_var, x_label, y_label, ymin, ymax, color_choice, n_highlight = 126, seed = 716) {
  set.seed(seed)
  highlight_ids <- sample(unique(data$CC.ID), n_highlight)
  data$highlight <- data$CC.ID %in% highlight_ids
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    geom_line(data = subset(data, !highlight), aes(group = CC.ID), alpha = 0.3, linewidth = 0.3, color = color_choice) +
    geom_line(data = subset(data, highlight), aes(group = CC.ID), alpha = 1, linewidth = 0.5, color = color_choice) +
    labs(x = x_label, y = y_label) +
    coord_cartesian(ylim = c(ymin, ymax)) +
    theme_minimal() +
    theme(
      axis.title.x = element_text(size = 24),
      axis.title.y = element_text(size = 24),
      axis.text = element_text(size = 16)
    )
}
plot_outliers <- function(data, x_var, y_var, x_label, y_label, ymin, ymax, outliers, color_choice) {
  outlier_CCIDs <- unique(outliers$CC.ID)
  data <- data %>%
    mutate(is_outlier = ifelse(CC.ID %in% outlier_CCIDs, TRUE, FALSE))
  non_outlier_data <- data %>%
    filter(!is_outlier)
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    geom_point(aes(color = is_outlier, alpha = is_outlier), size = 2, stroke = 1) +
    geom_line(aes(group = CC.ID), lty = 2, alpha = 1, color = "black") +
    geom_smooth(
      data = non_outlier_data,
      aes(x = .data[[x_var]], y = .data[[y_var]]),
      method = "lm", se = FALSE, size = 0.8, lty = 1, color = "black"
    ) +
    labs(x = x_label, y = y_label) +
    ylim(ymin, ymax) +
    scale_color_manual(
      values = c("TRUE" = "green", "FALSE" = color_choice),
      labels = c("TRUE" = "Outlier", "FALSE" = "Non-Outlier")
    ) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 1), guide = "none") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.text = element_text(size = 12), 
      axis.title.x = element_text(size = 16),
      axis.title.y = element_text(size = 16),
      axis.text.x = element_text(size = 10),
      axis.text.y = element_text(size = 10)
    ) +
    guides(color = guide_legend(title = NULL))
}
```
### Custom function for Mahalanobis distance 
```{r, include = TRUE}
label_mahalanobis_outliers <- function(data, id_col, variables, flag_col_name = "outlier_flag") {
  data_matrix <- data[, variables]
  dists <- mahalanobis(
    data_matrix,
    colMeans(data_matrix, na.rm = TRUE),
    cov(data_matrix, use = "complete.obs")
  )
  is_outlier <- dists > qchisq(0.975, df = length(variables))
  outlier_ids <- data[[id_col]][is_outlier]
  data[[flag_col_name]] <- ifelse(data[[id_col]] %in% outlier_ids, "Outlier", "Normal")
  return(data)
}
```

```{r}
sem_table <- function(fit_object, caption = "SEM Parameter Estimates") {
  params <- parameterEstimates(fit_object, standardized = TRUE) %>%
    filter(op %in% c("~", ":=")) %>%
    select(lhs, op, rhs, std.lv, se, z, pvalue)
  colnames(params) <- c("Outcome", "Op", "Predictor", "Std_Estimate", "SE", "Z", "p")
  params <- params %>%
    mutate(
      Std_Estimate = round(Std_Estimate, 2),
      SE = round(SE, 2),
      Z = round(Z, 2),
      p = round(p, 3),
      p_formatted = ifelse(p < 0.001, "<.001", sub("^0", "", formatC(p, format = "f", digits = 3)))
    )
  params <- params %>%
    mutate(
      Std_Estimate = ifelse(p < 0.05,
                            cell_spec(Std_Estimate, "html", bold = TRUE),
                            as.character(Std_Estimate)),
      p_formatted = ifelse(p < 0.05,
                           cell_spec(p_formatted, "html", bold = TRUE),
                           p_formatted)
    )
  params %>%
    select(Outcome, Op, Predictor, Std_Estimate, SE, Z, p_formatted) %>%
    rename(`p-value` = p_formatted) %>%
    kable("html", escape = FALSE, caption = caption) %>%
    kable_styling(full_width = FALSE, position = "center")
}
```

#### Read in data 
```{r}
mydata700     <- read.csv("data_preprocessed/CamCAN_cross_sectional_Cattell_GM_PGS_3Sep25.csv")
mydata_slopes <- read.csv("data_preprocessed/CamCAN_longitudinal_slopes_Cattell_GM_30Aug25.csv")
gm_G3         <- read.csv("data_preprocessed/CamCAN_longitudinal_GM_G3_4Sep25.csv")
```

### CROSS-SECTIONAL ANALYSIS 
#Descriptive statistic of the sample 
```{r}
cat("Cross-sectional age summary (A0_raw):\n")
print(summary(mydata700$A0_raw))
cat("Cross-sectional sex:\n")
print(table(factor(mydata700$sex, levels = c(1, 2), labels = c("Female", "Male"))))

cat("Longitudinal baseline age summary (A0_raw):\n")
print(summary(mydata_slopes$A0))
```

```{r}
gm_G3 <- gm_G3 %>% 
  mutate(mri_date = as.Date(mri_date)) %>%
  group_by(CC.ID) %>% 
  mutate(arm_id = any(wave == 0 & 
                      mri_date >= as.Date("2015-01-01") &
                      mri_date <= as.Date("2020-12-31"))) %>%
  ungroup() %>%
  arrange(CC.ID, mri_date)

p <- ggplot(gm_G3, aes(mri_date, dA, group = CC.ID)) +
  geom_line(color = "black", alpha = 0.05) +
  geom_point(aes(color = factor(wave), shape = arm_id), size = 1.5) +
  scale_color_manual(values = c("0"="green4","1"="purple3","2"="turquoise4"),
                     labels = c("Phase 2","Phase 3","Phase 5"),
                     name = NULL) +
  scale_shape_manual(values = c("FALSE"=16,"TRUE"=17),
                     labels = c("Arm 1","Arm 2"),
                     name = NULL) +
  scale_x_date(limits = as.Date(c("2010-01-01","2025-12-31")),
               breaks = seq(as.Date("2010-01-01"), as.Date("2025-01-01"), by = "5 years"),
               date_labels = "%Y") +
  labs(x = "MRI timeline", y = "Time since baseline (years)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_line(color = "grey90", linewidth = 0.1),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color = "grey90", linewidth = 0.1),
        panel.grid.minor.y = element_blank(),
        legend.text = element_text(size = 12),
        panel.ontop = FALSE,
        legend.position = "right") + 
  theme(plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "white", color = NA),
        legend.background = element_rect(fill = "white", color = NA))

ggsave("Figures/MRI_timeline.png", p, width = 8, height = 4, dpi = 300)
```

## Mediation models
### Outliers 
```{r}
vars_to_scale <- c("Age", "PGS", "Sex", "YoE", "Thickness", "Area")
mydata700[vars_to_scale] <- scale(mydata700[vars_to_scale])
mydata700 <- label_mahalanobis_outliers(
  data = mydata700,
  id_col = "CC.ID",
  variables = c("Cattell", "PGS", "Thickness", "Area", "Age", "Sex", "YoE"),
  flag_col_name = "outlier_flag"
)
all_outliers700 <- data.frame(CC.ID = mydata700$CC.ID[mydata700$outlier_flag == "Outlier"])
mydata700[mydata700$CC.ID %in% all_outliers700$CC.ID, setdiff(names(mydata700), "CC.ID")] <- NA
CamCAN_700 <- mydata700[!is.na(mydata700$Cattell), ]
```
#### Correlation matrix (final dataset)
```{r}
corr_vars <- CamCAN_700[, c("Age", "Cattell", "Thickness", "Area", "Sex", "PGS")]
colnames(corr_vars) <- c("Age","COG (Cattell)", "THK", "SA","PGS", "Sex")
matrix    <- cor(corr_vars)
testRes   <- cor.mtest(corr_vars, conf.level = 0.95)
png("Figures/cormatCamCAN700.png", width = 6, height = 6, units = "in", res = 600)
corrplot(matrix, method = 'number', p.mat = testRes$p, sig.level = 0.05,
         order = 'original', type = "lower", diag = FALSE, tl.col = "black")
dev.off()

corr_vars2 <- CamCAN_700[, c("Age", "Cattell", "Thickness", "Area", "Sex", "PGS", "YoE" )]
colnames(corr_vars2) <- c("Age","COG (Catell)", "THK", "SA","PGS", "YoE", "Sex")
matrix    <- cor(corr_vars2)
testRes   <- cor.mtest(corr_vars2, conf.level = 0.95)
png("Figures/cormatCamCAN700_YOE.png", width = 6, height = 6, units = "in", res = 600)
corrplot(matrix, method = 'number', p.mat = testRes$p, sig.level = 0.05,
         order = 'original', type = "lower", diag = FALSE, tl.col = "black")
dev.off()
```
### SEM1 (without YoE)
```{r}
SEM1 <- '
  Cattell ~ c1*Age + c2*PGS + b1*Thickness + b2*Area + c3*Sex
  
  Thickness  ~ a1*Age + a2*PGS        
  Area       ~ a3*Age + a4*PGS 
  Area       ~ Sex
  
  Thickness ~~ Area

  ATC := abs(a1 * b1)  
  ASC := abs(a3 * b2)  

  PTC := abs(a2 * b1)  
  PSC := abs(a4 * b2) 

  ABC := ATC + ASC     
  PBC := PTC + PSC     
  
  #Optional 
  Age_total := ABC + abs(c1)
  PGS_total := PBC + abs(c2)
  
  ATC_prop := ATC / Age_total
  ASC_prop := ASC / Age_total

  PTC_prop := PTC / PGS_total
  PSC_prop := PSC / PGS_total
'
SEM_fit1 <- sem(SEM1, data = CamCAN_700)
summary(SEM_fit1, fit.measures=TRUE, std = TRUE)
sem_table(SEM_fit1)

# custom_layout <- matrix(c(
#    0,  2,   
#   -1,  1,    
#    1,  1, 
#   -1,  0,  
#    1,  0,    
#    1.5,  1.7
# ), ncol = 2, byrow = TRUE) # [1] "Cattell" "Thickness" "Area" "YoE" "Age" "PGS" "Sex" 
# semPaths(SEM_fit1,what = "std", layout = custom_layout, edge.label.cex = 1.3, label.cex = 1,
#                       sizeMan = 10, residuals = FALSE, intercepts = FALSE, nCharNodes = 0, title = FALSE,
#                       mar = c(4, 4, 4, 4))

SEM1_equated_age <- '
  Cattell ~ c1*Age + c2*PGS + b1*Thickness + b1*Area + c3*Sex 

  Thickness ~ a1*Age + a2*PGS
  Area      ~ a1*Age + a4*PGS
  Area      ~ Sex
  
  Thickness ~~ Area 
  
  ATC := a1 * b1
  ASC := a1 * b1
  ABC := abs(ATC) + abs(ASC)
'
SEM_fit1_age_equated <- sem(SEM1_equated_age, data = CamCAN_700)
anova(SEM_fit1, SEM_fit1_age_equated)

SEM1_equated_PGS <- '
  Cattell ~ c1*Age + c2*PGS + b1*Thickness + b1*Area + c3*Sex

  Thickness ~ a1*Age + a2*PGS
  Area      ~ a3*Age + a2*PGS
  Area      ~ Sex
  
  Thickness ~~ Area 

  PTC := a2 * b1
  PSC := a2 * b1
  PBC := abs(PTC) + abs(PSC)
'
SEM_fit1_PGS_equated <- sem(SEM1_equated_PGS, data = CamCAN_700)
anova(SEM_fit1, SEM_fit1_PGS_equated)
```
### SEM PGS only (without YoE)
```{r}
SEM_PGS_only <- '
  Cattell ~ c1*PGS + b1*Thickness + b2*Area + c3*Sex

  Thickness ~ a1*PGS
  Area      ~ a2*PGS
  Area      ~ Sex

  Thickness ~~ Area

  PTC := a1 * b1
  PSC := a2 * b2
  PBC := abs(PTC) + abs(PSC)
'
PGS1_only <- sem(SEM_PGS_only, data = CamCAN_700)

SEM_PGS_only_equated <- '
  Cattell ~ c1*PGS + b1*Thickness + b1*Area + c3*Sex

  Thickness ~ a1*PGS
  Area      ~ a1*PGS
  Area      ~ Sex

  Thickness ~~ Area

  PTC := a1 * b1
  PSC := a1 * b1
  PBC := abs(PTC) + abs(PSC)
'
PGS1_only_equated <- sem(SEM_PGS_only_equated, data = CamCAN_700)
anova(PGS1_only, PGS1_only_equated)
```

### SEM2 (with YoE)
```{r}
SEM2 <- '
  Cattell ~ c1*Age + c2*PGS + b1*Thickness + b2*Area + c3*Sex + c4*YoE

  Thickness ~ a1*Age + a2*PGS        
  Area      ~ a3*Age + a4*PGS 
  
  Area      ~ Sex
  YoE       ~ Area

  Thickness ~~ Area

  ATC := abs(a1 * b1)  
  ASC := abs(a3 * b2)  

  PTC := abs(a2 * b1)  
  PSC := abs(a4 * b2) 

  ABC := ATC + ASC     
  PBC := PTC + PSC     

  Age_total := ABC + abs(c1)
  PGS_total := PBC + abs(c2)
  
  ATC_prop := ATC / Age_total
  ASC_prop := ASC / Age_total

  PTC_prop := PTC / PGS_total
  PSC_prop := PSC / PGS_total

'
SEM_fit2 <- sem(SEM2, data = CamCAN_700)
summary(SEM_fit2, fit.measures=TRUE, std = TRUE)
sem_table(SEM_fit2)

# custom_layout2 <- matrix(c(
#    0,  2,   
#   -1,  1,    
#    1,  1, 
#    1.6, 2,
#   -1,  0,  
#    1,  0,    
#    1.6,  1.4
# ), ncol = 2, byrow = TRUE) # [1] "Cattell" "Thickness" "Area" "Age" "PGS" "Sex" 
# 
# semPaths(SEM_fit2,what = "std", layout = custom_layout2, edge.label.cex = 1.3, label.cex = 1,
#                       sizeMan = 10, residuals = FALSE, intercepts = FALSE, nCharNodes = 0, title = FALSE,
#                       mar = c(4, 4, 4, 4))

SEM2_equated_age <- '
  Cattell ~ c1*Age + c2*PGS + b1*Thickness + b1*Area + c3*Sex + c4*YoE

  Thickness ~ a1*Age + a2*PGS
  Area      ~ a1*Age + a4*PGS
  
  Area      ~ Sex
  YoE       ~ Area
  
  Thickness ~~ Area 
  
  ATC := a1 * b1
  ASC := a1 * b1
  ABC := abs(ATC) + abs(ASC)
'
SEM_fit2_age_equated    <- sem(SEM2_equated_age, data = CamCAN_700)
anova(SEM_fit2, SEM_fit2_age_equated)
```

### LONGITUDINAL ANALYSIS 
#### Outliers 
```{r}
mydata_slopes <- read.csv("data_preprocessed/CamCAN_longitudinal_slopes_Cattell_GM_30Aug25.csv")

plotcross(mydata_slopes, "Thickness_A0_slope", "Cattell_slope_A0", "GM Thickness Change Rate", "Cognition Change Rate", min(mydata_slopes$Cattell_slope_A0), max(mydata_slopes$Cattell_slope_A0), color_choice = "darkorange")
plotcross(mydata_slopes, "Area_A0_slope", "Cattell_slope_A0", "GM Area Change Rate", "Cognition Change Rate", min(mydata_slopes$Cattell_slope_A0), max(mydata_slopes$Cattell_slope_A0),  color_choice = "turquoise4")

mydata_slopes$A0_raw <- mydata_slopes$A0
mydata_slopes$YoE    <- mydata_slopes$YoE_raw
mydata_slopes$Sex    <- scale(as.numeric(mydata_slopes$sex))

vars_to_scale <- c("Sex","YoE", "A0", "Cattell_slope_A0", "Thickness_A0_slope", "Area_A0_slope")
mydata_slopes[c("Sex", "YoE", "A0", "Cattell_slope_A0", "Thickness_A0_slope", "Area_A0_slope")] <- scale(mydata_slopes[vars_to_scale])

mydata_slopes <- label_mahalanobis_outliers(data = mydata_slopes, id_col = "CC.ID",
                 variables = c("Area_A0_slope", "Thickness_A0_slope", "Cattell_slope_A0", "Sex", "YoE", "A0"),
                 flag_col_name = "outlier_flag")

# mydata_slopes <- label_mahalanobis_outliers(
#   data      = mydata_slopes,
#   id_col    = "CC.ID",
#   variables = c("Area_A0dA_slope", "Thickness_A0dA_slope", "Cattell_slope_A0dA",
#                 #"WMV_slope",
#                 "Sex", "YoE",
#                 "A0"),
#
#   flag_col_name = "outlier_flag"
# )

# mydata_slopes <- label_mahalanobis_outliers(
#   data      = mydata_slopes,
#   id_col    = "CC.ID",
#   variables = c("Area_gamm_slope", "Thickness_gamm_slope", "Cattell_slope_gamm",
#                 #"WMV_slope",
#                 "Sex", "YoE",
#                 "A0"),
#
#   flag_col_name = "outlier_flag"
# )

all_outliers <- data.frame(CC.ID = mydata_slopes$CC.ID[mydata_slopes$outlier_flag == "Outlier"])
# all_outliers2 <- data.frame(CC.ID = mydata_slopes$CC.ID[mydata_slopes$outlier_flag == "Outlier"])
# all_outliers3 <- data.frame(CC.ID = mydata_slopes$CC.ID[mydata_slopes$outlier_flag == "Outlier"])
# all_outliers <- c("221585", "320321", "520076", "520542", "610392", "621020", "720718")

mydata_slopes$CC.ID    <- as.character(mydata_slopes$CC.ID)
outlier_thickness_plot <- plot_outliers(mydata_slopes, "Thickness_A0_slope", "Cattell_slope_A0", 
                                        "Thickness Change Rate", "Cattell Change Rate", 
                                         min(mydata_slopes$Cattell_slope_A0), 
                                         max(mydata_slopes$Cattell_slope_A0), 
                                         all_outliers, "darkorange")
outlier_thickness_plot

outlier_area_plot      <- plot_outliers(mydata_slopes, "Area_A0_slope", "Cattell_slope_A0", 
                                        "Surface Area Change Rate", "Cattell Change Rate", 
                                        min(mydata_slopes$Cattell_slope_A0), 
                                        max(mydata_slopes$Cattell_slope_A0), 
                                        all_outliers, "turquoise4")
outlier_area_plot
grid.arrange(outlier_thickness_plot, outlier_area_plot, ncol = 2)

mydata_slopes[mydata_slopes$CC.ID %in% all_outliers$CC.ID, setdiff(names(mydata_slopes), "CC.ID")] <- NA
CamCAN_slopes <- mydata_slopes[!is.na(mydata_slopes$Cattell_slope_A0), ]
```


## Univariate analysis of change in MR measures 
```{r}
# gm_G3$Thickness <- scale(gm_G3$Thickness)
# gm_G3$Area      <- scale(gm_G3$Area)

gm_G3[gm_G3$CC.ID %in% all_outliers$CC.ID, setdiff(names(gm_G3), "CC.ID")] <- NA
gm_G3 <- gm_G3[!is.na(gm_G3$MR_Age), ]

# thickness_change_plot <- plotit(gm_G3, "MR_Age", "Thickness", 
#                                 "Age", "THK", 
#                                 min(gm_G3$Thickness) , max(gm_G3$Thickness), "darkorange")
# 
# area_change_plot      <- plotit(gm_G3, "MR_Age", "Area", 
#                                 "Age", "SA", 
#                                  (min(gm_G3$Area)), (max(gm_G3$Area)), "turquoise4")
# 
# grid.arrange(thickness_change_plot, area_change_plot, ncol = 2)

thickness_change_plot2 <- plotit2(gm_G3, "MR_Age", "Thickness", 
                                "Age", "THK (mm)", 
                                min(gm_G3$Thickness) , max(gm_G3$Thickness), "darkorange")

area_change_plot2      <- plotit2(gm_G3, "MR_Age", "Area", 
                                "Age", "SA (mm²)", 
                                 (min(gm_G3$Area)), (max(gm_G3$Area)), "turquoise4")
png("Figures/THK_SA_age_CamCAN.png", width = 16, height = 8, units = "in", res = 300)
grid.arrange(thickness_change_plot2, area_change_plot2, ncol = 2)
dev.off()

gm_long <- gm_G3 %>%
  select(CC.ID, MR_Age, Thickness, Area) %>%
  pivot_longer(cols = c(Thickness, Area),names_to = "measure", values_to = "value")

gm_long <- gm_long %>%
  group_by(measure) %>%
  mutate(value_z = as.numeric(scale(value))) %>% ungroup()

ggplot(gm_long, aes(MR_Age, value_z, color = measure)) +
  geom_line(aes(group = interaction(CC.ID, measure)), alpha = 0.2, linewidth = 0.3) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE, linewidth = 1.4) +
  scale_color_manual(values = c("Thickness" = "darkorange","Area" = "turquoise4")) +
  labs(x = "Age", y = "z-score", title = NULL, color = NULL) +
  coord_cartesian(xlim = c(20, 100)) +
  theme_classic(base_size = 20) + 
  theme(legend.position = "none")
ggsave(filename = "Figures/GAMM_age_CamCAN.png", width = 8, height = 8, units = "in", dpi = 300)
```

###Univariate mixed-effects models
```{r}
Interaction <- (gm_G3$dA - mean(gm_G3$dA)) * (gm_G3$A0 - mean(gm_G3$A0))
OrthInteraction <- residuals(lmer(Interaction ~ dA + A0 + (1|CC.ID), data = gm_G3))

thickness_m <- lmer(Thickness ~ A0 + dA + OrthInteraction + (1|CC.ID), data=gm_G3)
area_m      <- lmer(Area      ~ A0 + dA + OrthInteraction + (1|CC.ID), data=gm_G3)

standardize_parameters(thickness_m)
standardize_parameters(area_m)

extract_lmer_summary <- function(model, model_name) {
  sm  <- summary(model)$coefficients
  std <- standardize_parameters(model)
  eta <- eta_squared(model, partial = TRUE)

  as.data.frame(sm) %>%
    rownames_to_column("Predictor") %>%
    filter(Predictor %in% c("A0", "dA", "OrthInteraction")) %>%
    left_join(
      std %>%
        filter(Parameter %in% c("A0", "dA", "OrthInteraction")) %>%
        select(Parameter, Std_Coefficient),
      by = c("Predictor" = "Parameter")
    ) %>%
    left_join(
      eta %>%
        filter(Parameter %in% c("A0", "dA", "OrthInteraction")) %>%
        select(Parameter, Eta2_partial),
      by = c("Predictor" = "Parameter")
    ) %>%
    transmute(
      Model = model_name,
      Predictor,
      beta      = round(Estimate, 4),
      beta_std  = round(Std_Coefficient, 3),
      eta2_p    = round(Eta2_partial, 3),
      p         = sub("^0\\.", ".", sprintf("%.3f", `Pr(>|t|)`))
    )
}
final_table <- bind_rows(
  extract_lmer_summary(thickness_m, "Thickness"),
  extract_lmer_summary(area_m,      "Area")
)
final_table %>%
  gt() %>%
  tab_header(
    title = "Trajectories age-related",
    subtitle = "Thickness, Area"
  ) %>%
  fmt_markdown(everything()) %>%
  opt_table_outline() %>%
  opt_all_caps()
```

## "Change - Change" analysis
```{r}
num_cols <- sapply(CamCAN_slopes, is.numeric) & !grepl("_raw$|^CC.ID$|^sex$", names(CamCAN_slopes))
CamCAN_slopes[num_cols] <- lapply(CamCAN_slopes[num_cols], scale)


outlier_rm_thickness_plot <- plotcross(CamCAN_slopes, "Thickness_A0_slope", "Cattell_slope_A0", 
                                       "THK change rate", "COG (Cattell) change rate", 
                                       min(CamCAN_slopes$Cattell_slope_A0), 
                                       max(CamCAN_slopes$Cattell_slope_A0), 
                                       "darkorange")

outlier_rm_area_plot      <- plotcross(CamCAN_slopes, "Area_A0_slope", "Cattell_slope_A0", 
                                       "SA change rate", "COG (Cattell) change rate", 
                                       min(CamCAN_slopes$Cattell_slope_A0), 
                                       max(CamCAN_slopes$Cattell_slope_A0), 
                                       "turquoise4")

png("Figures/change_change_CamCAN.png", width = 16, height = 8, units = "in", res = 300)
grid.arrange(outlier_rm_thickness_plot, outlier_rm_area_plot, ncol = 2)
dev.off()
```

```{r}
adjustments <- c("A0", "A0dA", "gamm")
slope_cols <- grepl("slope", names(CamCAN_slopes))

CamCAN_slopes[slope_cols] <- scale(CamCAN_slopes[slope_cols])

model_grid <- expand.grid(adj = adjustments, 
                          measure = c("Thickness", "Area"),
                          stringsAsFactors = FALSE)

models_simple <- lapply(seq_len(nrow(model_grid)), function(i) {
  adj <- model_grid$adj[i]
  m   <- model_grid$measure[i]
  yvar <- paste0("Cattell_slope_", adj)
  xvar <- paste0(m, "_", adj, "_slope")
  fml <- as.formula(paste(yvar, "~", xvar))
  lm(fml, data = CamCAN_slopes)
})
names(models_simple) <- paste0("m_", model_grid$measure, "_", model_grid$adj)

results_simple <- do.call(
  rbind,
  lapply(names(models_simple), function(nm) {

    m  <- models_simple[[nm]]
    sm <- summary(m)$coefficients

    eta2 <- summary(m)$r.squared  # correct for simple lm

    data.frame(
      model     = nm,
      predictor = rownames(sm),
      beta      = sm[, "Estimate"],
      p         = sm[, "Pr(>|t|)"],
      eta2      = c(NA, eta2),  # intercept gets NA
      row.names = NULL
    )
  })
) %>%
  dplyr::filter(predictor != "(Intercept)")


final_table <- results_simple %>%
  mutate(
    Adjustment = case_when(
      grepl("_A0$", model)   ~ "A0",
      grepl("_A0dA$", model) ~ "A0dA",
      grepl("_gamm$", model) ~ "GAMM"
    ),
    Measure = case_when(
      grepl("^m_Thickness", model) ~ "Thickness",
      grepl("^m_Area", model)      ~ "Area"
    ),
    stat = sprintf("%.2f, η² = %.2f, p = %.3f", beta, eta2, p),
    stat = sub("p = 0\\.", "p = .", stat)
  ) %>%
  select(Adjustment, Measure, stat) %>%
  pivot_wider(names_from = Measure, values_from = stat) %>%
  arrange(factor(Adjustment, levels = c("A0", "A0dA", "GAMM")))

steiger_table <- tibble(
  Adjustment = c("A0", "A0dA", "GAMM"),
  adj_suffix = c("A0", "A0dA", "gamm")
) %>%
  rowwise() %>% mutate(
    steiger = list({
      y  <- CamCAN_slopes[[paste0("Cattell_slope_", adj_suffix)]]
      x1 <- CamCAN_slopes[[paste0("Thickness_", adj_suffix, "_slope")]]
      x2 <- CamCAN_slopes[[paste0("Area_", adj_suffix, "_slope")]]
      ok <- complete.cases(y, x1, x2)
      ct <- cocor.dep.groups.overlap(
        r.jk = cor(y[ok], x1[ok]),
        r.jh = cor(y[ok], x2[ok]),
        r.kh = cor(x1[ok], x2[ok]),
        n    = sum(ok), alternative = "two.sided")
      tibble(
        z_steiger = sprintf("%.2f", ct@steiger1980$statistic),
        p_steiger = sub("^0\\.", ".", sprintf("%.3f", ct@steiger1980$p.value)),
        n = sum(ok))
    })
  ) %>% unnest(steiger) %>% select(-adj_suffix)
final_table_all <- final_table %>% left_join(steiger_table, by = "Adjustment")
final_table_all %>%
  gt() %>%
  tab_header(
    title = "Model Results",
    subtitle = "Thickness, Area, and Steiger Comparisons"
  ) %>%
  fmt_markdown(everything()) %>%
  opt_table_outline() %>%
  opt_all_caps()
### SANITY CHECK!
# summary(lm(Cattell_slope_A0dA ~ Thickness_A0dA_slope, data = CamCAN_slopes))
# summary(lm(Cattell_slope_A0dA ~ Area_A0dA_slope, data = CamCAN_slopes))
```

```{r}
roi_cols <- grep("^mean_(?!MeanThickness_thickness|WhiteSurfArea_area).*_A0_slope$",
                 names(CamCAN_slopes),value=TRUE,perl=TRUE)
roi_meta <- data.frame(col=roi_cols,
            region  = sub("^mean_(.*)_(thickness|area)_A0_slope$", "\\1",roi_cols),
            measure = sub("^mean_.*_(thickness|area)_A0_slope$", "\\1", roi_cols),
            stringsAsFactors = FALSE)
roi_long <- do.call(rbind, lapply(seq_len(nrow(roi_meta)), function(i) {
  cl <- roi_meta$col[i]
  df <- data.frame(
    ROI     = CamCAN_slopes[[cl]],
    Cattell = CamCAN_slopes$Cattell_slope_A0
  )
  df <- df[complete.cases(df), ]
  sm <- summary(lm(Cattell ~ ROI, data = df))$coefficients["ROI", ]
  data.frame(
    region  = roi_meta$region[i],
    measure = roi_meta$measure[i],
    beta    = round(sm["Estimate"], 2),
    p       = sm["Pr(>|t|)"],   
    n       = nrow(df),
    stringsAsFactors = FALSE
  )
}))

roi_long$p_fdr <- ave(roi_long$p, roi_long$measure, FUN = function(x) p.adjust(x, method = "fdr"))
roi_table <- reshape(roi_long, idvar = "region",timevar = "measure", direction = "wide")
p_cols <- grep("^p(\\.|_fdr\\.)", names(roi_table), value = TRUE)
roi_table[p_cols] <- lapply(roi_table[p_cols],function(x) sub("^0\\.", ".", sprintf("%.3f", x)))
roi_table %>%
  gt() %>%
  tab_header(
    title = "ROI Model Results",
    subtitle = "Thickness, Area by ROI"
  ) %>%
  fmt_markdown(everything()) %>%
  opt_table_outline() %>%
  opt_all_caps()

### SANITY CHECK !
# summary(lm(Cattell_slope_A0 ~ mean_middletemporal_area_A0_slope, data = CamCAN_slopes))

roi_long$sig <- roi_long$p_fdr < 0.05
roi_long$region <- factor(
  roi_long$region,
  levels = rev(unique(roi_long$region))
)
ggplot(roi_long, aes(x = beta, y = region, color = measure, shape = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(position = position_dodge(width = 0.3), size = 2) +
  scale_shape_manual(values = c(`FALSE` = 1, `TRUE` = 16)) +
  scale_x_continuous(limits = c(-0.30, 0.30)) +
  scale_color_manual(values = c(thickness = "darkorange", area = "turquoise4")) +
  labs(title = "CamCAN", x = NULL, y = NULL) +
  theme_bw() +
  theme(panel.grid.major.y = element_blank(), legend.position = "none", plot.title = element_text(hjust = 0.5))
ggsave("Figures/roi_effects_long_CamCAN.png", width = 3, height = 9, dpi = 300)
```

```{r}
CamCAN_slopes$Thickness_baseline <- scale(CamCAN_slopes$Thickness_baseline)
CamCAN_slopes$Area_baseline <- scale(CamCAN_slopes$Area_baseline)

models_sb <- lapply(adjustments, function(adj) {
  yvar <- paste0("Cattell_slope_", adj)
  x_th <- paste0("Thickness_", adj, "_slope")
  x_ar <- paste0("Area_", adj, "_slope")
  b_th <- paste0("Thickness_baseline")
  b_ar <- paste0("Area_baseline")
  fml <- as.formula(
    paste(yvar, "~", x_th, "+", x_ar, "+", b_th, "+", b_ar)
  )
  lm(fml, data = CamCAN_slopes)
})
names(models_sb) <- paste0("m2_", adjustments)

results <- do.call(
  rbind,
  lapply(names(models_sb), function(nm) {
    m  <- models_sb[[nm]]
    sm <- summary(m)$coefficients
    eta <- eta_squared(m, partial = TRUE)
    as.data.frame(sm) %>%
      tibble::rownames_to_column("predictor") %>%
      left_join(
        eta %>%
          select(Parameter, Eta2_partial),
        by = c("predictor" = "Parameter")
      ) %>%
      transmute(
        model     = nm,
        predictor = predictor,
        beta      = Estimate,
        p         = `Pr(>|t|)`,
        eta2      = Eta2_partial
      )
  })
)

results <- results %>% filter(grepl("Thickness_.*_slope|Area_.*_slope|Thickness_baseline|Area_baseline", predictor))

wide_models <- results %>%
  mutate(
    Adjustment = case_when(
      grepl("^m2_A0$", model)   ~ "A0",
      grepl("^m2_A0dA$", model) ~ "A0dA",
      grepl("^m2_gamm$", model) ~ "GAMM"
    ),
    Predictor_clean = case_when(
      grepl("Thickness_.*_slope", predictor) ~ "Thickness_slope",
      grepl("Area_.*_slope", predictor)      ~ "Area_slope",
      grepl("Thickness_baseline", predictor) ~ "Thickness_baseline",
      grepl("Area_baseline", predictor)      ~ "Area_baseline"
    ),
    beta_r = round(beta, 2),
    eta_r  = round(eta2, 2),
    p_r    = round(p, 3),
    stat   = paste0(beta_r, ", η² = ", eta_r, ", p = ", sub("^0\\.", ".", p_r))
  ) %>%
  select(Adjustment, Predictor_clean, stat) %>%
  pivot_wider(
    names_from  = Predictor_clean,
    values_from = stat
  ) %>%
  arrange(factor(Adjustment, levels = c("A0", "A0dA", "GAMM"))) %>%
  select(
    Adjustment,
    Thickness_slope,
    Area_slope,
    Thickness_baseline,
    Area_baseline
  )

extract_lh <- function(mod, hyp) {
  ct <- as.data.frame(linearHypothesis(mod, hyp))
  row <- ct[2, , drop = FALSE]
  
  stat_col <- intersect(c("Chisq", "F"), names(row))[1]
  p_col    <- intersect(c("Pr(>Chisq)", "Pr(>F)"), names(row))[1]
  
  stat_val <- suppressWarnings(as.numeric(row[[stat_col]]))
  p_val    <- suppressWarnings(as.numeric(row[[p_col]]))
  tibble(stat_type = stat_col, stat = round(stat_val, 2), p_lh = sub("^0\\.", ".", sprintf("%.3f", p_val)))
}

lh_table <- tibble(
  Adjustment = c("A0", "A0dA", "GAMM"),
  model_name = c("m2_A0", "m2_A0dA", "m2_gamm"),
  contrast   = c(
    "Thickness_A0_slope - Area_A0_slope = 0",
    "Thickness_A0dA_slope - Area_A0dA_slope = 0",
    "Thickness_gamm_slope - Area_gamm_slope = 0"
  )
) %>%
  mutate(out = pmap(list(model_name, contrast), \(mn, h) extract_lh(models_sb[[mn]], h))) %>%
  tidyr::unnest(out) %>%
  select(-model_name, -contrast)

final_models_table <- wide_models %>% left_join(lh_table, by = "Adjustment")
final_models_table %>%
  gt() %>%
  tab_header(
    title = "Slope + Baseline Results",
    subtitle = "Thickness, Area"
  ) %>%
  fmt_markdown(everything()) %>%
  opt_table_outline() %>%
  opt_all_caps()
```


## Longitudinal SEM (BSCMs models)
```{r}
BLCS_thickness <-'
Thickness2 ~ 1*Thickness0 #YES
dThickness =~ 1*Thickness2 #YES
Thickness2 ~ 0*1 #YES
Thickness2 ~~ 0*Thickness2 #YES

Cattell2 ~ 1*Cattell0 #YES
dCattell =~ 1*Cattell2 #YES
dCattell ~ 1 #YES
Cattell0 ~  1 #YES
Cattell2 ~ 0*1 #YES

dCattell ~~   dCattell #YES
Cattell0 ~~   Cattell0 #YES
Cattell2 ~~ 0*Cattell2 #YES

dThickness ~ 1 #YES
Thickness0 ~ 1 #YES
dThickness ~~ dThickness #YES
Thickness0 ~~ Thickness0 #YES

dThickness~Cattell0+Thickness0 #YES
dCattell~Thickness0+Cattell0 #YES

Cattell0 ~~  Thickness0 #YES
dCattell ~~ dThickness #YES
'

fitBLCS_thickness <- lavaan(BLCS_thickness, data=CamCAN_slopes, estimator='mlr',fixed.x=TRUE,missing='fiml')
summary(fitBLCS_thickness, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)

thickness_plot <- semPlot::semPaths(fitBLCS_thickness,
                  whatLabels = "std",
                  layout = "tree",
                  rotation = 2,
                  edge.label.cex = 1.5,
                  label.cex = 1.2,
                  sizeMan = 9,
                  sizeLat = 9,
                  nCharNodes = 3,
                  intercepts = FALSE,
                  residuals = FALSE,
                  edge.color = "black")

parameterEstimates(fitBLCS_thickness, standardized = FALSE) |>
  subset(op == "~" & rhs %in% c("Thickness0","Cattell0"))

thickness_plot <- mark_sig(thickness_plot, fitBLCS_thickness)
graphics::plot(thickness_plot)

BLCS_thickness_age <-'
Thickness2 ~ 1*Thickness0 #YES
dThickness =~ 1*Thickness2 #YES
Thickness2 ~ 0*1 #YES
Thickness2 ~~ 0*Thickness2 #YES

Cattell2 ~ 1*Cattell0 #YES
dCattell =~ 1*Cattell2 #YES
dCattell ~ 1 #YES
Cattell0 ~  1 #YES
Cattell2 ~ 0*1 #YES

dCattell ~~   dCattell #YES
Cattell0 ~~   Cattell0 #YES
Cattell2 ~~ 0*Cattell2 #YES

dThickness ~ 1 #YES
Thickness0 ~ 1 #YES
dThickness ~~ dThickness #YES
Thickness0 ~~ Thickness0 #YES

dThickness~Cattell0+Thickness0 #YES
dCattell~Thickness0+Cattell0 #YES

Cattell0 ~~  Thickness0 #YES
dCattell ~~ dThickness #YES

dThickness ~ A0 
dCattell   ~ A0 
'

fitBLCS_thickness_age <- lavaan(BLCS_thickness_age, data=CamCAN_slopes, estimator='mlr',fixed.x=TRUE,missing='fiml')
summary(fitBLCS_thickness_age, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)

```

```{r}
BLCS_area <-'

Area2 ~ 1*Area0
dArea =~ 1*Area2
Area2 ~ 0*1
Area2 ~~ 0*Area2

Cattell2 ~ 1*Cattell0
dCattell =~ 1*Cattell2
dCattell ~ 1
Cattell0 ~  1
Cattell2 ~ 0*1

dCattell ~~  dCattell
Cattell0 ~~   Cattell0
Cattell2 ~~ 0*Cattell2

dArea ~ 1
Area0 ~ 1
dArea ~~ dArea
Area0 ~~ Area0

dArea~Cattell0+Area0
dCattell~Area0+Cattell0

Cattell0 ~~ Area0
dCattell ~~ dArea

'
fitBLCS_area <- lavaan(BLCS_area, data=CamCAN_slopes, estimator='mlr',fixed.x=TRUE,missing='fiml')
summary(fitBLCS_area, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)

area_plot <- semPlot::semPaths(fitBLCS_area,
                  whatLabels = "std",
                  layout = "tree",
                  rotation = 2,
                  edge.label.cex = 1.5,
                  label.cex = 1.2,
                  sizeMan = 9,
                  sizeLat = 9,
                  nCharNodes = 3,
                  intercepts = FALSE,
                  residuals = FALSE,
                  edge.color = "black")

area_plot <- mark_sig(area_plot, fitBLCS_area)
graphics::plot(area_plot)

BLCS_area_age <-'

Area2 ~ 1*Area0
dArea =~ 1*Area2
Area2 ~ 0*1
Area2 ~~ 0*Area2

Cattell2 ~ 1*Cattell0
dCattell =~ 1*Cattell2
dCattell ~ 1
Cattell0 ~  1
Cattell2 ~ 0*1

dCattell ~~  dCattell
Cattell0 ~~   Cattell0
Cattell2 ~~ 0*Cattell2

dArea ~ 1
Area0 ~ 1
dArea ~~ dArea
Area0 ~~ Area0

dArea~Cattell0+Area0
dCattell~Area0+Cattell0

Cattell0 ~~  Area0
dCattell ~~ dArea

dArea    ~ A0 
dCattell ~ A0 
'
fitBLCS_area_age <- lavaan(BLCS_area_age, data=CamCAN_slopes, estimator='mlr',fixed.x=TRUE,missing='fiml')
summary(fitBLCS_area_age, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)

```
