---
title: "2.1_data_preprocessing_OASIS"
output: html_document
date: "2025-11-23"
---

```{r}
library(tidyverse)
library(lmerTest)
library(psych)
library(data.table)
library(miceadds)
library(patchwork)
library(gamm4)
```

# Longitudinal data preparation 
```{r}
freesurf <- read.csv("data_raw/OASIS3_Freesurfer_output.csv")
cog      <- read.csv("data_raw/OASIS3_UDSc1_cognitive_assessments.csv")
sex      <- read.csv("data_raw/OASIS_Sex_DOB.csv")
edu      <- read.csv("data_raw/OASIS_Edu_Others.csv")

vol_cols <- grep("_volume$", names(freesurf), value = TRUE)
thickness_cols <- grep("_thickness$", names(freesurf), value = TRUE)

freesurf$id <- freesurf$Subject
cog$id      <- cog$OASISID
sex$id      <- sex$Subject

cog <- merge(cog, edu[, c("OASISID", "EDUC")], by = "OASISID", all.x = TRUE)
cog <- merge(cog, sex[, c("id", "M.F")], by = "id", all.x = TRUE)

names(cog)[names(cog) == "EDUC"] <- "YoE"
names(cog)[names(cog) == "M.F"]  <- "Sex"

cog$Sex      <- ifelse(cog$Sex == "M", 1, ifelse(cog$Sex == "F", 2, NA))
cog$sex_code <- cog$Sex

freesurf$lag <- as.numeric(sub(".*OAS\\d+_MR_d(\\d+)", "\\1", freesurf$MR_session))
cog$lag      <- as.numeric(sub(".*OAS\\d+_psychometrics_d(\\d+)", "\\1", cog$OASIS_session_label))
```

### Initial assessment
```{r}
length(unique(freesurf$id))
length(unique(cog$id))
```

```{r}
freesurf$dx   <- NA  
freesurf$wave <- NA

cog$A0   <- NA  
cog$dx   <- NA  
cog$wave <- NA

freesurf$lag_years <- round(freesurf$lag/365, 2)
cog$lag_years      <- round(cog$lag/365, 2)

for (i in unique(freesurf$Subject)) {
  rows <- which(freesurf$Subject == i)
  lags <- freesurf$lag[rows]
  if (all(is.na(lags))) next
  freesurf$dx[rows]   <- round((lags - min(lags, na.rm = TRUE))/365, 2)
  freesurf$wave[rows] <- order(order(freesurf$dx[rows]))-1
}

#Same for cognition 
for (i in unique(cog$OASISID)) {
  rows <- which(cog$OASISID == i)
  lags <- cog$lag[rows]
  if (all(is.na(lags))) next
  cog$dx[rows] <- round((lags - min(lags, na.rm = TRUE))/365, 2)
  cog$A0[rows] <- round(min(cog$age.at.visit[rows]) - min(cog$lag_years[rows]), 2) 
  #additonal step: calculate absolute A0 at the study start
  cog$wave[rows] <- order(order(cog$dx[rows]))-1
}

#Calculate actual age at MR scans and actual age at Cog assess 
cog$cog_age          <- as.numeric(cog$age.at.visit) #actual cog assess age 
cog$dx_unscaled      <- cog$dx 
freesurf$dx_unscaled <- freesurf$dx 

freesurf$A0     <- cog$A0[match(freesurf$Subject, cog$OASISID)]
freesurf$MR_age <- freesurf$A0 + freesurf$lag_years 
#actual MRI scan age (absolute A0 + lag_years per each scan)

#Pull out age at baseline for cog and age at baseline for MR (Not the same as A0!)
cog <- cog %>%
  group_by(id) %>%
  mutate(cog_age0 = first(cog_age[dx == 0])) %>%
  ungroup()

freesurf <- freesurf %>%
  group_by(id) %>%
  mutate(MR_age0 = first(MR_age[dx == 0])) %>%
  ungroup()

# Calculate how many waves each ID has  
# cog$n_waves      <- as.numeric(ave(cog$id, cog$id, FUN = length))
# freesurf$n_waves <- as.numeric(ave(freesurf$id, freesurf$id, FUN = length))

length(unique(freesurf$id[freesurf$dx > 6]))
length(unique(cog$id[cog$dx > 6]))

length(unique(freesurf$id[freesurf$dx > 2]))
length(unique(cog$id[cog$dx > 2]))
```
*A0* - absolute age in the beginning of the study (does not correspond to any assessment and should not be used in modelling)
*lag_years* - actual years since the absolute beginning of the study (do not use for any modelling)
*MR_age0* - individual age at baseline for MR scan (repeated for waves in each ID, use to model cross-sectional effects of age in brain trajectories)
*cog_age0* - individual age at baseline for cog assess (repeated in each ID, use to model cross-sectional effects of age in cognitive trajectories)
*dx* - standardised lags between assessments (set to 0 for the first assessment) (use to model change effects)
*n_waves* - how many waves each ID has (do not use for modelling, descriptive statistic)

```{r}
plot_df <- bind_rows(
  freesurf %>% mutate(modality = "MRI"),
  cog %>% mutate(modality = "COG")
) %>%
  filter(!is.na(lag_years), !is.na(dx))

plot_study_timeline <- ggplot(plot_df, aes(lag_years, dx)) +
  geom_hex(aes(fill = modality), alpha = 0.9) +
  geom_segment(
    data = data.frame(x = 0, y = 0, xend = 35, yend = 35),
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,
    linetype = "dashed", color = "grey40", linewidth = 0.4,
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  geom_segment(
    data = data.frame(intercept = c(-5, -10, -15, -20)),
    aes(x = -intercept, y = 0, xend = 35, yend = 35 + intercept),
    inherit.aes = FALSE,
    linetype = "dotted", color = "grey70", linewidth = 0.4,
    arrow = arrow(length = unit(0.25, "cm"), type = "open")
  ) +
  facet_wrap(~ modality) +
  scale_fill_manual(values = c("COG" = "#D8B4FE", "MRI" = "lightblue1")) +
  labs(
    x = "Time of recruitment since the study start (years)",
    y = "Time between assessments (years)",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )

ggsave("Figures/study_timeline_OASIS.png", plot = plot_study_timeline, width = 8, height = 4, dpi = 300, bg = "white")
```

```{r}
make_raster_clean <- function(df, id_col, session_col, dx_col, title, fill_col) {
  dfv <- df %>%
    filter(!is.na(.data[[dx_col]])) %>%
    distinct(.data[[id_col]], .data[[session_col]], .data[[dx_col]]) %>%
    rename(id = all_of(id_col), dx = all_of(dx_col))
  id_order <- dfv %>%
    group_by(id) %>%
    summarise(max_dx = max(dx), .groups = "drop") %>%
    arrange(desc(max_dx)) %>%
    mutate(id_ord = row_number())
  grid_df <- expand_grid(
    id = id_order$id,
    dx_bin = 0:30
  ) %>%
    left_join(id_order, by = "id") %>%
    left_join(dfv %>% mutate(dx_bin = floor(dx)), by = c("id", "dx_bin")) %>%
    mutate(present = !is.na(dx))
  ggplot(grid_df, aes(dx_bin, id_ord)) +
    geom_tile(aes(fill = present), height = 0.9, width = 0.9) +
    scale_fill_manual(values = c("TRUE" = fill_col, "FALSE" = "white")) +
    scale_x_continuous(limits = c(0, 30), expand = c(0, 0)) +
    scale_y_reverse(expand = c(0, 0)) +
    labs(
      title = title,
      x = "Time since indvidual baseline (years)",
      y = "Participants"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      legend.position = "none",
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 8))
    )
}

p_cog <- make_raster_clean(cog, id_col = "id",session_col = "OASIS_session_label", dx_col = "dx", title = "COG", fill_col = "#D8B4FE")
p_mri <- make_raster_clean(freesurf,id_col = "id",session_col = "MR_session",dx_col = "dx",title = "MRI",fill_col = "lightblue1")

ggsave("Figures/Dropout_wave_pattern_cog.png", p_cog, width = 4, height = 4, dpi = 600, bg = "white")
ggsave("Figures/Dropout_wave_pattern_mri.png", p_mri, width = 4, height = 4, dpi = 600, bg = "white")

png("Figures/Dropout_wave_pattern.png", width = 8, height = 4, units = "in", res = 600)
grid.arrange(p_cog, p_mri, ncol = 2)
dev.off()
```

## Cognitive variables 
```{r}
cog_vars <- c("ANIMALS", "VEG", "SIM", "PSY021", 
              "digsym",  "digback", "tma", "tmb", 
              "lettnum", "slosson", "switch", "block")
sapply(cog[cog_vars], function(x) round(100 - mean(!is.na(x)) * 100, 1)) #check missing data % 

# ANIMALS   VEG    SIM    PSY021  digsym  digback  tma    tmb    lettnum  slosson  switch   block 
#    98.1   77.9   28.0   76.9    78.4    60.3     91.1   89.1   67.4     1.2      62.3     11.8 

# VARS to keep (less 30% missing): ANIMALS, VEG, PSY021, digsym, tma, tmb 
```

## PCA  
```{r}

cog$tmb[cog$tmb >= 180] <- NA
cog$tma[cog$tma >= 150] <- NA

# plot(density(cog$tmb, na.rm = TRUE))
# plot(density(cog$tma, na.rm = TRUE))
# plot(density(cog$digsym, na.rm = TRUE))
# plot(density(cog$PSY021, na.rm = TRUE))

cog$PSY021  <- scale((as.numeric(cog$PSY021)))
cog$ANIMALS <- scale(as.numeric(cog$ANIMALS))
cog$VEG     <- scale(as.numeric(cog$VEG))
cog$digsym  <- scale((as.numeric(cog$digsym)))
cog$digback <- scale((as.numeric(cog$digback)))
cog$lettnum <- scale((as.numeric(cog$lettnum)))
cog$switch  <- scale((as.numeric(cog$switch)))
cog$tma     <- (-1)*scale(log(as.numeric(cog$tma)))
cog$tmb     <- (-1)*scale(log(as.numeric(cog$tmb)))

# plot(density(cog$tmb, na.rm = TRUE))
# plot(density(cog$tma, na.rm = TRUE))
# plot(density(cog$digsym, na.rm = TRUE))
# plot(density(cog$PSY021, na.rm = TRUE))

cog      <- cog %>% arrange(OASISID, dx)
cog$prac <- 1
cog$prac[!duplicated(cog$OASISID)] <- 0
cog$prac <- as.factor(cog$prac)

```

```{r}

# pca_vars <- c("ANIMALS", "VEG", "digsym", "tma", "tmb", "PSY021")
# cog[pca_vars] <- lapply(cog[pca_vars], function(x) as.numeric(as.vector(x)))
# 
# vars_for_mice <- c("id", "dx", "Sex", "YoE", "prac", pca_vars)
# data_mice <- cog[, vars_for_mice]
# data_mice$id <- as.integer(as.factor(data_mice$id))
# 
# ini  <- mice(data_mice, maxit = 0)
# pred <- ini$predictorMatrix
# meth <- ini$method
# 
# pca_rows <- match(pca_vars, colnames(pred))
# id_col   <- match("id", colnames(pred))
# dx_col   <- match("dx", colnames(pred))
# sex_col  <- match("sex", colnames(pred))
# yoe_col  <- match("YoE", colnames(pred))
# prac_col  <- match("prac", colnames(pred))
# 
# pred[pca_rows, id_col]  <- -2   # cluster variable
# pred[pca_rows, dx_col]  <-  1   # include dx as predictor
# pred[pca_rows, sex_col] <-  1   # include sex as predictor
# pred[pca_rows, yoe_col] <-  1   # include YoE as predictor
# pred[pca_rows, prac_col] <-  1   # include YoE as predictor
# 
# # Do not impute id, dx, sex, or YoE
# pred[id_col, ]  <- 0
# pred[dx_col, ]  <- 0
# pred[sex_col, ] <- 0
# pred[yoe_col, ] <- 0
# pred[prac_col, ] <- 0
# 
# # Set method for PCA variables (multilevel predictive mean matching)
# meth[pca_rows] <- "2l.pmm"
# 
# # Run MICE
# imp <- mice(data_mice, method = meth, predictorMatrix = pred,
#             m = 30, maxit = 10, seed = 123)
# 
# # Extract completed dataset
# cog_pca_imputed <- complete(imp, 1)
# cog_pca_imputed$id <- data_mice$id
# cog[pca_vars] <- cog_pca_imputed[pca_vars]
# 
# write.csv(cog, "data_preprocessed/OASIS_longitudinal_imputed.csv", row.names = FALSE)
```

```{r}
cog <- read.csv("data_preprocessed/OASIS_longitudinal_imputed.csv")

pca_vars <- c("ANIMALS", "VEG", "digsym", "tma", "tmb", "PSY021")
cog$dA <- cog$cog_age - cog$cog_age0
min_dx_rows <- do.call(rbind, by(cog, cog$id, function(df) df[which.min(df$dx), ]))
cog_pca_data <- na.omit(min_dx_rows[, pca_vars])

pca_cog <- prcomp(cog_pca_data, center = TRUE, scale. = TRUE)
pc1_loadings_cog <- -pca_cog$rotation[, 1]

all_cog   <- as.matrix(cog[, pca_vars])
cog$gCog0 <- as.numeric(all_cog %*% pc1_loadings_cog)

for (v in pca_vars) {
  f <- as.formula(paste0(v, " ~ prac + (1|id)"))
  idx <- complete.cases(cog[, c(v, "cog_age0", "prac", "id")])
  m <- lmer(f, data = cog[idx, ], REML = TRUE)
  cog[[paste0(v, "_A0")]] <- NA
  cog[[paste0(v, "_A0")]][idx] <- residuals(m)
}
cog$gCog_A0 <- as.numeric(
  as.matrix(cog[, paste0(pca_vars, "_A0")]) %*% pc1_loadings_cog
)

for (v in pca_vars) {
  f <- as.formula(paste0(v, " ~ cog_age0:dA + prac + (1|id)"))
  idx <- complete.cases(cog[, c(v, "cog_age0", "dA", "prac", "id")])
  m <- lmer(f, data = cog[idx, ], REML = TRUE)
  cog[[paste0(v, "_A0dA")]] <- NA
  cog[[paste0(v, "_A0dA")]][idx] <- residuals(m)
}
cog$gCog_A0dA <- as.numeric(
  as.matrix(cog[, paste0(pca_vars, "_A0dA")]) %*% pc1_loadings_cog
)

for (v in pca_vars) {
  f <- as.formula(paste0(v, " ~ s(cog_age) + prac"))
  idx <- complete.cases(cog[, c(v, "cog_age", "prac", "id")])
  m <- gamm(f, data   = cog[idx, ], random = list(id = ~1), REML   = TRUE)
  cog[[paste0(v, "_gamm")]] <- NA
  cog[[paste0(v, "_gamm")]][idx] <-
    residuals(m$gam, type = "response")
}
cog$gCog_gamm <- as.numeric(
  as.matrix(cog[, paste0(pca_vars, "_gamm")]) %*% pc1_loadings_cog
)

pc_summary    <- summary(pca_cog)
var_explained <- round(pc_summary$importance[2, 1:3] * 100, 2)
pc1_loadings  <- round(pca_cog$rotation[, 1]^2 * 100, 2)
pc1_animals   <- pc1_loadings["ANIMALS"]
pc1_veg       <- pc1_loadings["VEG"]
pc1_digsym    <- pc1_loadings["digsym"]
pc1_tma       <- pc1_loadings["tma"]
pc1_tmb       <- pc1_loadings["tmb"]
pc1_psy021    <- pc1_loadings["PSY021"]

pc1_animals
pc1_veg
pc1_digsym
pc1_tma
pc1_tmb
pc1_psy021

```

- DSST: `r pc1_digsym`%  
- Trail Making B: `r pc1_tmb`%  
- Trail Making A: `r pc1_tma`%  
- WAIS Block Design: `r pc1_psy021`%  
- Category Fluency (Animals): `r pc1_animals`%  
- Category Fluency (Vegetables): `r pc1_veg`%

### Cogntive slopes 
```{r}
eligible_ids <- unique(cog$id[cog$dx_unscaled >= 2])
cog_long     <- cog[cog$id %in% eligible_ids, ]

filtered_cog <- cog_long[complete.cases(cog_long[, c("dx", "gCog_A0", "gCog_A0dA", "gCog_gamm", "gCog0")]),]

unique_ids <- unique(filtered_cog$id)
n          <- length(unique_ids)

gCog_slope_A0    <- rep(NA, n)
gCog_slope_A0dA  <- rep(NA, n)
gCog_slope_gamm  <- rep(NA, n)

Sex <- rep(NA, n)
YoE <- rep(NA, n)

for (i in seq_len(n)) {

  this_id     <- unique_ids[i]
  subset_data <- filtered_cog[filtered_cog$id == this_id, ]

  if (nrow(subset_data) >= 2) {

    m_A0 <- lm(gCog_A0 ~ dx, data = subset_data)
    gCog_slope_A0[i] <- coef(m_A0)[["dx"]]

    m_A0dA <- lm(gCog_A0dA ~ dx, data = subset_data)
    gCog_slope_A0dA[i] <- coef(m_A0dA)[["dx"]]

    m_gammm <- lm(gCog_gamm ~ dx, data = subset_data)
    gCog_slope_gamm[i] <- coef(m_gammm)[["dx"]]

    wave0  <- subset_data[which.min(subset_data$dx), ]
    Sex[i] <- wave0$Sex[1]
    YoE[i] <- wave0$YoE[1]
  }
}

cog_slopes <- data.frame(
  id               = unique_ids,
  gCog_slope_A0    = gCog_slope_A0,
  gCog_slope_A0dA  = gCog_slope_A0dA,
  gCog_slope_gamm  = gCog_slope_gamm,
  Sex              = Sex,
  YoE              = YoE,
  stringsAsFactors = FALSE
)

m_re <- lmer(gCog0 ~ scale(A0) + scale(A0):dx + scale(prac) + (dx|id), data = filtered_cog, REML = TRUE)
cog_slopes$gCog_slope_ranef <- ranef(m_re)$id[, "dx"]
```

### Brain slopes 
```{r}
freesurf$Volume    <- rowMeans(freesurf[, vol_cols], na.rm = TRUE)
freesurf$Thickness <- rowMeans(freesurf[, thickness_cols], na.rm = TRUE)
freesurf$Area      <- (freesurf$L.SurfArea + freesurf$R.SurfArea) / 2

freesurf <- freesurf[
  !is.na(freesurf$Thickness) &
  !is.na(freesurf$Area) &
  !is.na(freesurf$MR_age0),
]

freesurf$dA <- freesurf$MR_age - freesurf$MR_age0

eligible_ids  <- unique(freesurf$id[freesurf$dx >= 2])
freesurf_long <- freesurf %>% filter(id %in% eligible_ids)

unique_ids <- unique(freesurf_long$id)
n <- length(unique_ids)

lh_thickness_cols <- grep("^lh_.*_thickness$", names(freesurf_long), value = TRUE)
rh_thickness_cols <- grep("^rh_.*_thickness$", names(freesurf_long), value = TRUE)

freesurf_long[paste0("mean_", sub("^lh_", "", lh_thickness_cols))] <-
  (freesurf_long[lh_thickness_cols] + freesurf_long[rh_thickness_cols]) / 2

ROIs_thickness_cols <- grep("^mean_.*_thickness", names(freesurf_long), value = TRUE)
roi_names <- sub("^mean_(.*)_thickness$", "\\1", ROIs_thickness_cols)

brain_vars <- c("Thickness", "Area", "CorticalWhiteMatterVol")

for (v in brain_vars) {

  m_A0 <- lmer(as.formula(paste0(v, " ~ 1 + (1|id)")), data = freesurf_long, REML = TRUE)
  freesurf_long[[paste0(v, "_A0")]] <- residuals(m_A0)

  m_A0dA <- lmer(as.formula(paste0(v, " ~ MR_age0:dA + (1|id)")), data = freesurf_long,REML = TRUE)
  freesurf_long[[paste0(v, "_A0dA")]] <- residuals(m_A0dA)

  m_gamm <- gamm(as.formula(paste0(v, " ~ s(MR_age)")),data   = freesurf_long, random = list(id = ~1), REML   = TRUE)
  freesurf_long[[paste0(v, "_gamm")]] <- residuals(m_gamm$gam, type = "response")
}

for (roi in ROIs_thickness_cols) {
  
  m_A0 <- lmer(as.formula(paste0(roi, " ~ 1 + (1|id)")),data = freesurf_long, REML = TRUE)
  freesurf_long[[paste0(roi, "_A0")]] <- residuals(m_A0)
  
  m_A0dA <- lmer(as.formula(paste0(roi, " ~ MR_age0:dA + (1|id)")), data = freesurf_long, REML = TRUE)
  freesurf_long[[paste0(roi, "_A0dA")]] <- residuals(m_A0dA)

  m_gamm <- gamm(as.formula(paste0(roi, " ~ s(MR_age)")),data   = freesurf_long, random = list(id = ~1),REML   = TRUE)
  freesurf_long[[paste0(roi, "_gamm")]] <- residuals(m_gamm$gam, type = "response")
}

mk_slope_mat <- function(suffix, vars) {
  matrix(
    NA, nrow = n, ncol = length(vars),
    dimnames = list(NULL, paste0(vars, "_", suffix, "_slope"))
  )
}

brain_slopes_A0    <- mk_slope_mat("A0", brain_vars)
brain_slopes_A0dA  <- mk_slope_mat("A0dA", brain_vars)
brain_slopes_gamm  <- mk_slope_mat("gamm", brain_vars)

roi_slopes_A0      <- mk_slope_mat("A0", roi_names)
roi_slopes_A0dA    <- mk_slope_mat("A0dA", roi_names)
roi_slopes_gamm    <- mk_slope_mat("gamm", roi_names)

Thickness_baseline <- numeric(n)
Area_baseline      <- numeric(n)

A0 <- numeric(n)

for (i in seq_len(n)) {

  sid <- unique_ids[i]
  d   <- freesurf_long[freesurf_long$id == sid, ]

  if (nrow(d) >= 2) {
    
    for (j in seq_along(brain_vars)) {
      v <- brain_vars[j]
      brain_slopes_A0[i, j]   <- coef(lm(d[[paste0(v,"_A0")]]   ~ d$dx))[2]
      brain_slopes_A0dA[i, j] <- coef(lm(d[[paste0(v,"_A0dA")]] ~ d$dx))[2]
      brain_slopes_gamm[i, j] <- coef(lm(d[[paste0(v,"_gamm")]] ~ d$dx))[2]
    }

    for (j in seq_along(roi_names)) {
      roi <- paste0("mean_", roi_names[j], "_thickness")
      roi_slopes_A0[i, j]   <- coef(lm(d[[paste0(roi,"_A0")]]   ~ d$dx))[2]
      roi_slopes_A0dA[i, j] <- coef(lm(d[[paste0(roi,"_A0dA")]] ~ d$dx))[2]
      roi_slopes_gamm[i, j] <- coef(lm(d[[paste0(roi,"_gamm")]] ~ d$dx))[2]
    }

    wave0 <- d[which.min(d$dx), ]
    A0[i]  <- wave0$MR_age0
    Thickness_baseline[i] <- wave0$Thickness[1]
    Area_baseline[i]      <- wave0$Area[1]
  }
}

gm_slopes <- data.frame(
  id = unique_ids,
  A0 = A0,
  Thickness_baseline = Thickness_baseline,
  Area_baseline = Area_baseline, 
  brain_slopes_A0,
  brain_slopes_A0dA,
  brain_slopes_gamm,
  roi_slopes_A0,
  roi_slopes_A0dA,
  roi_slopes_gamm,
  stringsAsFactors = FALSE
)

m_re_thk <- lmer(Thickness ~ scale(MR_age0) + scale(MR_age0):dx + (dx|id), data = freesurf_long, REML = TRUE)
gm_slopes$Thickness_ranef_slope <- ranef(m_re_thk)$id[, "dx"]

m_re_sa <- lmer(Area ~ scale(MR_age0) + scale(MR_age0):dx + (dx|id), data = freesurf_long, REML = TRUE)
gm_slopes$Area_ranef_slope <- ranef(m_re_sa)$id[, "dx"]
```

```{r}

mydata_slopes <- merge(gm_slopes, cog_slopes, by = "id")

length(unique(mydata_slopes$id))
min(mydata_slopes$A0)
max(mydata_slopes$A0)
mean(mydata_slopes$A0)
sd(mydata_slopes$A0)

length(unique(mydata_slopes$id[mydata_slopes$Sex == 1]))
length(unique(mydata_slopes$id[mydata_slopes$Sex == 2]))

```

```{r}

freesurf_long_wav <- freesurf_long[freesurf_long$id %in% mydata_slopes$id, ]
cog_long_wav      <- cog_long[cog_long$id %in% mydata_slopes$id, ]

# Prepare data for MR barplot
wav_df_mr <- as.data.frame(table(table(freesurf_long_wav$id)))
colnames(wav_df_mr) <- c("NumSessions", "NumParticipants")
wav_df_mr$NumSessions <- as.numeric(as.character(wav_df_mr$NumSessions))
wav_df_mr$NumParticipants <- as.numeric(wav_df_mr$NumParticipants)

a <- ggplot(wav_df_mr, aes(x = NumSessions, y = NumParticipants)) +
  geom_bar(stat = "identity", fill = "darkgrey", color = "white") +
  geom_text(aes(label = NumParticipants), vjust = -0.5, size = 5) +
  labs(x = "Number of MR Testing Sessions", y = "Number of Participants") +
  scale_x_continuous(breaks = c(2, 4, 6, 8, 10)) +
  theme_minimal()

# Prepare data for Cognitive barplot
wav_df_cog <- as.data.frame(table(table(cog_long_wav$id)))
colnames(wav_df_cog) <- c("NumSessions", "NumParticipants")
wav_df_cog$NumSessions <- as.numeric(as.character(wav_df_cog$NumSessions))
wav_df_cog$NumParticipants <- as.numeric(wav_df_cog$NumParticipants)

b <- ggplot(wav_df_cog, aes(x = NumSessions, y = NumParticipants)) +
  geom_bar(stat = "identity", fill = "darkgrey", color = "white") +
  geom_text(aes(label = NumParticipants), vjust = -0.5, size = 5) +
  labs(x = "Number of Cognitive Testing Sessions", y = "Number of Participants") +
  ylim(0, max(wav_df_cog$NumParticipants) * 1.15) +
  theme_minimal()


c <- ggplot(freesurf_long_wav, aes(x = dx, y = factor(id))) +  
     geom_point(alpha = 0.7, size = 1) +       
     geom_line(aes(group = id), alpha = 0.3) + 
     # geom_vline(xintercept = 6, linetype = 5, color = "yellow", size = 1) +
     scale_y_discrete(expand = expansion(mult = c(0.1, 0.1))) +  # add margin top & bottom
     theme_minimal() +
     labs(x = "Time (MR data)", y = "Participants") +
     theme(
      legend.position = "none",
      axis.text.y = element_blank(),   
      axis.ticks.y = element_blank() 
     ) + xlim(0,30)

cog_long_wav <- cog_long[cog_long$id %in% mydata_slopes$id, ]
d <- ggplot(cog_long_wav, aes(x = dx, y = factor(id))) +  
     geom_point(alpha = 0.7, size = 1) +       
     geom_line(aes(group = id), alpha = 0.3) + 
     # geom_vline(xintercept = 6, linetype = 5, color = "yellow", size = 1) +
     scale_y_discrete(expand = expansion(mult = c(0.1, 0.1))) +  # add margin top & bottom
     theme_minimal() +
     labs(x = "Time (Cognitive data)", y = "Participants") +
     theme(
      legend.position = "none",
      axis.text.y = element_blank(),   
      axis.ticks.y = element_blank() 
     ) + xlim(0,30)


label_theme <- theme(
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text.x  = element_text(size = 14)
)


a <- a + label_theme
b <- b + label_theme
c <- c + label_theme
d <- d + label_theme

combined <- (a | b) / (c | d)
print(combined)

ggsave("combined_plot.png", plot = combined, width = 16, height = 12, dpi = 300)
```

```{r}
write.csv(mydata_slopes, "data_preprocessed/OASIS_longitudinal_gCog_GM_5Sep25.csv", row.names = FALSE)
write.csv(freesurf_long, "data_preprocessed/OASIS_longitudinal_GM_5Sep25.csv", row.names = FALSE)
```

# Cross-sectional data preparation 
### Lag-matching algorithm 
```{r}
fs <- as.data.table(freesurf)
cg <- as.data.table(cog)

fs[, lag_years_fs  := lag_years]
cg[, lag_years_cog := lag_years]

setnames(fs, old = c("id", "dx", "wave", "MR_age"), new = c("id_fs", "dx_fs", "wave_fs", "MR_age_fs"), skip_absent = TRUE)
setnames(cg, old = c("id", "dx", "wave", "cog_age"), new = c("id_cg", "dx_cg", "wave_cg", "cog_age_cg"), skip_absent = TRUE)

results <- vector("list", nrow(fs))
cg[, matched := FALSE]

for (i in seq_len(nrow(fs))) {

  id_i  <- fs$id_fs[i]
  lag_i <- fs$lag_years_fs[i]

  candidates <- cg[id_cg == id_i & matched == FALSE]

  if (nrow(candidates) == 0) {
    results[[i]] <- cbind(fs[i], cog_match = NA, lag_diff = NA)
    next
  }

  diffs <- abs(candidates$lag_years_cog - lag_i)
  min_idx <- which.min(diffs)
  best_match <- candidates[min_idx]

  best_match$lag_diff <- diffs[min_idx]
  results[[i]] <- cbind(fs[i], best_match)

  cg[rownames(cg) == rownames(best_match), matched := TRUE]
}

mydata_matched            <- rbindlist(results, fill = TRUE)
mydata_different_rates    <- mydata_matched[(dx_fs != 0) | (dx_fs == 0 & lag_diff < 1)]
mydata_corresponding_lags <- mydata_matched[mydata_matched$lag_diff < 1, ]

length(unique(mydata_matched$id_fs)) #1316
length(unique(mydata_different_rates$id_fs)) #1275
length(unique(mydata_corresponding_lags$id_fs)) #1269

mydata                    <- mydata_corresponding_lags
mydata_different_rates    <- as_tibble(mydata_different_rates, .name_repair = "unique")
mydata_different_rates$id <- mydata_different_rates$id_fs

mydata_cross <- mydata_different_rates

mydata_cross <- do.call(rbind, by(mydata_cross, mydata_cross$id, function(df) {
  df[which.min(df$dx_fs), , drop = FALSE]
}))

mydata_cross$Age <- mydata_cross$A0...210
mydata_cross     <- mydata_cross[, c("id", "Age", "gCog0", "Thickness", "Area", "Sex", "YoE")]
mydata_cross     <- na.omit(mydata_cross[, c("id", "Age", "gCog0", "Thickness", "Area", "Sex", "YoE")])
```

```{r}
write.csv(mydata_cross, "data_preprocessed/OASIS_cross_sectional_gCog_GM_5Sep25.csv", row.names = FALSE)
write.csv(mydata, "data_preprocessed/OASIS_corresponding_lags.csv", row.names = FALSE)
```
